---
title: "r_int_day_3_data_transformation"
author: "Nicholus Tint Zaw"
date: "2022-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(dslabs)
```



# Content 

This lecture focus on the implementation of data wrangling using the `dplyer` package and will cover some exercise using the popular functions from `dplyer` package. 

1. filter
1. arrange
1. select
1. mutate
1. summarise
1. group_by



## Some useful commands for exploring dataset 


```{r df, results = 'hide'}
# load dataset 

df <- polls_us_election_2016 # from dslabs package

head(df) # print first 6 obs
tail(df) # print last 6 obs 

names(df) # variable/column names

unique(df$grade) # inspect unique values in a specific variable/column

typeof(df$grade) # inspect the type of variable 
class(df$grade) # inspect the type of variable 

table(df$grade) # frequency table

```



## `filter` : select the sub-set of the dataset 

```{r filter, results ='hide'}

filter(df, grade == "D")
filter(df, state == "Ohio")

# filter with multiple conditions 
interested_grades <- c("D", "B")

filter(df, grade %in% interested_grades) # one line

filter(df, grade == "D" | grade == "B")

```
Using `filter` with pipe `(%>%)` function. 

```{r filterpipe , results ='hide'}

df %>%
  filter(state == "Illinois")


iris %>% filter(Sepal.Length < mean(Sepal.Length))

```



## `arrange` : order the dataset by given variable

```{r  arrange , results ='hide'}

head(arrange(df, samplesize)) # ascending order - default  

head(arrange(df,desc(population))) # descending order 


head(arrange(df, population))
head(arrange(df, desc(population)))

```

use with pipe function,

```{r arrangepipe , results ='hide'}

df %>% arrange(population)

df %>% arrange(desc(population))

```



## `select`: keep only variables require for data processing 

```{r select , results ='hide'}

head(select(df, state, samplesize, population))


head(select(df, -(c(state, population, samplesize))))

```

```{r selecthelper, results ='hide'}

# starts_with("")

df %>%
  select(starts_with("adj"))

# ends_with("")
df %>% 
  select(ends_with("date"))

# contains("")
df %>%
  select(contains("raw"))

# matches("(.)\\1") - regular expression 
df %>% 
  select(matches("raw")) # check the result. 

# what is the different between matches and contains? 

df1 <- data.frame(colnm = 1:5, col1 = 24, col2 = 46)
df1 %>% 
    select(contains("col"))

df1 %>%
  select(matches("col\\d")) 

# num_range("x", 1:3)
# pls check at help file - type ?dplyr::select in console. 


```


Wanna study more about regular expression, check [here](https://paulvanderlaken.files.wordpress.com/2017/08/r-regular-expression-cheetsheat.pdf).




## `mutate`: adding new column to existing one

```{r mutate, results ='hide'}

df %>%
  mutate(ss_small = ifelse(samplesize < 1000, 1, 0)) %>%
  select(samplesize, ss_small)

# if want to add into existing dataframe, override the existing dataframe with resulted dataframe. 
df <- df %>% 
  mutate(ss_small = ifelse(samplesize < 1000, 1, 0)) # check the variable names and numbers 

```





## `summarise`: perform sumstat functions 

```{r sum, results ='hide'}
summarise(df, ss_mean = mean(samplesize)) # what happened!

summarize(df, ss_mean = mean(samplesize, na.rm = TRUE)) # Missing values are contagious

# Missing values are contagious
NA + 1

10 * 100 * NA

100/NA


# sumstat table for samplesize var
df %>%
  summarise(man = mean(samplesize, na.rm = TRUE), 
            sd = sd(samplesize, na.rm = TRUE), 
            median = median(samplesize, na.rm = TRUE),
            min = min(samplesize, na.rm = TRUE), 
            max = max(samplesize, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), 
               names_to  = "parameter")

```



## `group_by`: manipulation at group level, sub-group level 

```{r subgrp , results ='hide'}

# get the sample size mean value per state
df %>%
  group_by(state) %>%
  summarise(mean = mean(samplesize, na.rm = TRUE))


df %>%
  group_by(population) %>%
  summarise(mean = mean(samplesize, na.rm = TRUE))


df %>%
  group_by(population) %>%
  summarise(n = n())

df %>%
  group_by(state) %>%
  count()

```




## Exercise 

+ Use `iris` dataset and calculate sumstat by `Species`. 
+ Select the only observation from `versicolor` and `virginica` species. 



